Stack Safety is a Security Property

### 2021-04-07 tex/main.tex
Further edits based on Andrew's feedback

Since we rely so heavily on code annotations, we must ask: what if the annotations
are wrong? First, suppose a call or return occurs that is not annotated.
In this case, the answer is simple:
there has been no call or return\apt{from whose perspective?}, only an instruction resembling such an operation, and
therefore no associated change in permissions. This is particularly interesting in the
case of a false return, where control may appear to have returned to the caller, but
the enforcement mechanism must still protect the caller's memory.

Alternatively, suppose an annotated instruction is not really an actual call or return.
What happens in this case depends on the enforcement mechanism and the nature of the false
annotation. A given enforcement mechanism might place a static well-formedness criterion
on the code it applies to. For instance, our\apt{or \citeauthor{DBLP:conf/sp/RoesslerD18}'s ? }
micro-policies expect that call and return
annotations occur on jump instructions. On the other hand, not all jumps are proper
calls or returns, only those preceded by appropriate sequences: in these cases the
bad annotation might appear well formed but the system should either respect the change
in permissions or, if doing so would compromise security, fail-stop rather than
execute the annotated instruction.\apt{hard to follow}

@@

\apt{I'm unclear about how many of these assumptions are relevant to the properties
  as opposed to the (micro)policies.}

Our primary target is a RISC-V architecture extended with the PIPE architecture, but the
model makes very few assumptions about the machine it applies to. We assume a register
machine with a program counter \(\PCname\) and stack pointer \(\rsp\); other details of
the machine's ISA are unimportant.

Our definitions are largely parametric over the layout of the stack in memory, but for
purposes of explanation we will describe a standard layout. Memory consists of an
instruction region and a stack that grows upwards.\apt{if we want to be ``standard''
  we really should have the stack grown downwards.}
(For simplicity,
we do not consider the heap. Integrating a memory-safety property for the
heap in the style of \citet{DBLP:conf/post/AmorimHP18} is a natural extension
that we discuss in~\Cref{sec:relwork}.) The stack has a conventional structure, with
contiguously allocated per-function frames.

The boundary between the frames of a caller and callee is indicated by the $\rsp$ register
at the time of the call. The stack can be used to pass arguments or return results
in the usual way, by placing them at the top of the caller's frame. This will be modeled
as a form of sharing in section \sna{cite the sharing section}.
Callers and callees can also communicate via registers if they wish; indeed, our
definitions say nothing at all about registers (except $\PCname$ and $\rsp$).

\ifaftersubmission\bcp{This implicitly amounts to choosing a caller-saves register
  convention, though, doesn't it?\apt{I think it is neutral; we could add separate caller register protection if we wanted.}}\fi

@@

\begin{subfigure}{.4\textwidth}
\begin{verbatim}
int main() {
0:     int x = 42, y = 0;
1[C]:  y = f(10);
2:     print (x + y);
}

int f(int z) {
10:     print z;
11[R]:  return (z+z);
}
\end{verbatim}

\caption{A Simple Call}
\label{fig:simple-program}
\end{subfigure}
\begin{subfigure}{.4\textwidth}
\begin{center}
\begin{tabular}{l l}
{\tt 0} &
\memoryaddrs{4em}
\memory{2}{\unsealc}[{\makebox[0pt]{Unsealed}}]
~$\cdots$
\\
{\tt 1} &
\memoryaddrs{12em}
\memory{3}{\unsealc}
~$\cdots$
\MemoryLabel{-14em}{0.75em}{42}
\MemoryLabel{-10em}{0.75em}{0}
\vspace{.5em}
\\
{\tt 10} &
\memoryaddrs{17em}
\memory{2}{\mainsealc}[{\makebox[0pt]{Sealed(0)}}]%
\memory{2}{\unsealc}
~$\cdots$
\MemoryLabel{-10em}{0.75em}{2}
\\
{\tt 2} &
\memoryaddrs{12em}
\memory{3}{\unsealc}
~$\cdots$
\MemoryLabel{-14em}{0.75em}{42}
\MemoryLabel{-10em}{0.75em}{0}
\MemoryLabel{-7em}{0.75em}{2}
\\
\end{tabular}
\end{center}

\caption{Trace of Stack Domains}

\label{fig:simple-trace}
\end{subfigure}
\end{figure}

@@

\Cref{fig:simple-program} shows a simple example, written in C-like pseudo-code.
It consists of a {\tt main} function that performs a call to a function {\tt f},
which prints its argument, performs arithmetic, and returns a result.
\Cref{fig:simple-trace} shows the stack as it is divided into domains throughout
the program execution. We use high-level code for compactness, but reason about
a standard compilation of each function. \apt{Note assumption that arguments and return values are in registers.}
Line numbers correspond to the code
addresses just before the code that implements the line, and annotations
{\tt C} and {\tt R} indicate that an instruction is annotated as a call or
return, respectively.
\apt{In trace, clarify difference between values (like 42,0) and line numbers (like 2).}

We also assume that the machine is word- (not byte-) addressed and that the
machine communicates with the outside world through a single memory-mapped
output port whose address is elided (so a {\tt print} statement will be
compiled to a store to this location), and that the stack grows upwards.

In the initial memory layout the entire stack is currently {\em unsealed},
meaning that it is eligible to be written to. At the call from {\tt main} to
{\tt f}, {\tt main}'s locals (as designated \apt{``delimited''? in any case, how?} by the stack pointer) are
{\em sealed} and labeled with the depth of their owner, in this case 0.
Sealing is the fundamental integrity contract between the caller and callee:
{\tt main} expects that {\tt x} and {\tt y} will be unchanged when it gets control back.
This gives us our simple, intuitive statement of {\em stepwise integrity}:
if a component is sealed before a step, it will be unchanged afterward. Stepwise
refers to the fact that the property must hold on every step. Later we will
loosen this condition in the interest of extensionality -- the caller only
cares about changes it can see after the return, not changes during execution.

Suppose we insert an instruction after line 10 that stores 5 to the stack pointer - 1.\apt{Actually, -2.
  But this also raises the question of if/how return addresses are protected.}
Then we would be changing {\tt y}, which is sealed, in violation of stepwise integrity.

@@

int f() {
10:     int z = *($SP-1);
11[R]:  return (z+z);
}
\end{verbatim}
\label{fig:conf1-program}
\end{subfigure}
\begin{subfigure}{.59\textwidth}
\centering
\begin{tabular}{l l l}
  {\tt 0} &
  \multicolumn{2}{c}{
    \memoryaddrs{4em}
    \memory{3}{\unsealc}
    ~$\cdots$
    \MemoryLabel{-14em}{0.75em}{a}
    \MemoryLabel{-10em}{0.75em}{b}
    \MemoryLabel{-7em}{0.75em}{c}
    \vspace{.5em}
  } \\
  {\tt 1} &
  \multicolumn{2}{c}{
    \memoryaddrs{12em}
    \memory{3}{\unsealc}
    ~$\cdots$
    \MemoryLabel{-14em}{0.75em}{42}
    \MemoryLabel{-10em}{0.75em}{0}
    \MemoryLabel{-7em}{0.75em}{c}
    \vspace{.5em}
  } \\
  {\tt 10} &
  \memoryaddrs{20em}
  \memory{2}{\mainsealc}
  \memory{2}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{42}
  \MemoryLabel{-14em}{0.75em}{0}
  \MemoryLabel{-10em}{0.75em}{2}
  \MemoryLabel{-7em}{0.75em}{\bf d}
 &
  \memoryaddrs{20em}
  \memory{2}{\mainsealc}
  \memory{2}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{a'}
  \MemoryLabel{-14em}{0.75em}{b'}
  \MemoryLabel{-10em}{0.75em}{2}
  \MemoryLabel{-7em}{0.75em}{\bf d'}
  \\
  {\tt 11} &
  \memoryaddrs{20em}
  \memory{2}{\mainsealc}
  \memory{2}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{42}
  \MemoryLabel{-14em}{0.75em}{0}
  \MemoryLabel{-10em}{0.75em}{2}
  \MemoryLabel{-7em}{0.75em}{\bf 0}
  &
  \memoryaddrs{20em}
  \memory{2}{\mainsealc}
  \memory{2}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{a'}
  \MemoryLabel{-14em}{0.75em}{b'}
  \MemoryLabel{-10em}{0.75em}{2}
  \MemoryLabel{-7em}{0.75em}{\bf b'}
  \\
\end{tabular}

@@

\section{Extension to Cheri}

>>>>

Since we rely so heavily on code annotations, we must ask: what if the annotations
are wrong? First, suppose a call or return occurs that is not annotated.
In this case, the answer is simple: as far as the properties are concerned, there has
been no call or return, only an instruction sequence resembling such an operation, and
therefore no associated change in permissions. This is particularly interesting in the
case of a false return, where control may appear to have returned to the caller, but
the enforcement mechanism must still protect the caller's memory.

Alternatively, an annotated instruction might not really be a call or return.
In the context of \citeauthor{DBLP:conf/sp/RoesslerD18}'s micro-policies, well-behaved calls
and returns are identified with ``blessed sequences'' of instructions, which the tag policy
will force to be executed from start to finish. The annotated instruction should always
be the final instruction of the sequence. What if an annotated instruction exists outside
such a sequence? We will statically rule out some scenarios with a well-formedness criterion
on initial states: annotated instructions will initially be tagged apropriately, as the
final instruction in the call or return sequence. This is all we need for the policy to
protect itself: it can then enforce that the call or return only occurs if that instruction
is a jump that follows the blessed sequence. So, even in our strong attacker model, we need
very little in the way of extra guarantees.

In general, regardless of enforcement mechanism, we take this approach: 
a given enforcement mechanism may place a static well-formedness criterion
on the code it applies to. Weaker enforcement mechanisms might also need dynamic
reasonableness conditions to prevent the defender from sabotaging enforcement, as
discussed in section \ref{sec:cheri}.

@@

Our properties are defined in terms of an abstract machine model that makes very few
assumptions about the machine that instantiates it.  We assume a register
machine with a program counter \(\PCname\) and stack pointer \(\rsp\); other
details of the machine's ISA are unimportant. We assume that we know ahead of time
the layout of memory, including code (with annotations) and which addresses are within
the bounds of the stack.
%
We further parameterize over the structure of the stack, particularly the means by
which a caller identifies the data that should be protected during a call. We term
this the {\em sealing convention} and model it as a relation between machine states
and addresses. Then from any call, those the sealing convention relates to the
state will be protected until the return.

@@

\begin{subfigure}{.4\textwidth}
\begin{verbatim}
int main() {
0:     int x = 42, y = 0;
1[C]:  y = f(10);
2[R]:  return (x + y);
}

int f(int z) {
10:     $(1,0) = 0;
11[R]:  return (z+z);
}
\end{verbatim}
\end{subfigure}
\begin{subfigure}{.4\textwidth}
\begin{center}
\begin{tabular}{l l}
{\tt 0} &
\memory{2}{\unsealc}[{\makebox[0pt]{Unsealed}}]
\vspace{.5em}
\\
{\tt 1} &
\memory{2}{\unsealc}
\MemoryLabel{-6em}{0.75em}{42}
\MemoryLabel{-4em}{0.75em}{0}
\vspace{.5em}
\\
{\tt 10} &
\memory[1em]{1}{\emptyoutc}
\hspace{.5em}
\memory{2}{\mainsealc}[{\makebox[0pt]{Sealed(0)}}]%
\\
{\tt 11} &
\memory[1em]{1}{\emptyoutc}
\hspace{.5em}
\memory{2}{\mainsealc}[{\makebox[0pt]{Sealed(0)}}]%
\MemoryLabel{-6em}{0.75em}{\bf 0}
\end{tabular}
\end{center}

\vspace{\abovedisplayskip}
\end{subfigure}
\label{fig:int1}
\caption{Integrity Violation}
\end{figure}

@@

Our definitions are largely parametric over the layout of the stack in memory, so we
will give our initial examples in a simplified layout. Treat the stack as a non-empty
list of frames, each a contiguous block of memory (that may be of length zero). When
the active function allocates local variables, it does so by increasing the size of the
head of the stack; variables declared in nested scopes increase the size further when
entering the scope, and decrease it when leaving. During a call, the sealing convention
designates the entire head frame to be sealed, and a new frame is pushed onto the stack.
At a return, the stack is popped. We omit return addresses from these examples, as they
focus on local state encapsulation.

We have described an idealized stack, so we need to add a way for the attacker to violate
stack safety. We define the expression {\tt \$(i,j)} as accessing the \(i\)th frame in the
list at the offset \(j\). We will show how uses of this expression produce violations
of both integrity and confidentiality. Besides this operator, examples are in C-like
pseudocode. Line numbers represent the point in execution before executing all instructions
compiled from that line, with annotations in brackets. (Initially, only {\tt C} for call
and {\tt R} for return.) Variables are already allocated at the first line of a block.
All arguments and return values are passed in registers.

%The boundary between the frames of a caller and callee is indicated by the $\rsp$ register
%at the time of the call. The stack can be used to pass arguments or return results
%in the usual way, by placing them at the top of the caller's frame. This will be modeled
%as a form of sharing in section \sna{cite the sharing section}.
%Callers and callees can also communicate via registers if they wish; indeed, our
%definitions say nothing at all about registers (except $\PCname$ and $\rsp$).

%\ifaftersubmission\bcp{This implicitly amounts to choosing a caller-saves register
%  convention, though, doesn't it?\apt{I think it is neutral; we could add separate caller register protection if we wanted.}}\fi

\Cref{fig:int1} shows an integrity violation. It consists of a {\tt main}
function that performs a call to a function {\tt f}, which overwrites {\tt main}'s
variable {\tt x}, performs arithmetic, and returns a result.
In the initial memory layout the only frame is currently {\em unsealed},
meaning that it is eligible to be written to. At the call from {\tt main} to
{\tt f}, {\tt main}'s locals are {\em sealed} and labeled with the depth of their owner,
in this case 0. Sealing is the fundamental integrity contract between the caller and callee:
{\tt main} expects that {\tt x} and {\tt y} will be unchanged when it gets control back.
This gives us our simple, intuitive statement of {\em stepwise integrity}:
if a component is sealed before a step, it will be unchanged afterward. Stepwise
refers to the fact that the property must hold on every step. %Later we will
%loosen this condition in the interest of extensionality -- the caller only
%cares about changes it can see after the return, not changes during execution.
In this case, after line 10, {\tt f} has violated integrity by writing to {\tt x}.

\sna{I'm experimenting with the example style, and it ends here currently, so below
  this is the old version in which the stack is contiguous. Current plan is to switch
  to contiguous stack when adding stack passing.}

@@

int f() {
10:     int z = *($SP-1);
11[R]:  return (z+z);
}
\end{verbatim}
\label{fig:conf1-program}
\end{subfigure}
\begin{subfigure}{.59\textwidth}
\centering
\begin{tabular}{l l l}
  {\tt 0} &
  \multicolumn{2}{c}{
    \memory{3}{\unsealc}
    ~$\cdots$
    \MemoryLabel{-14em}{0.75em}{a}
    \MemoryLabel{-10em}{0.75em}{b}
    \MemoryLabel{-7em}{0.75em}{c}
    \vspace{.5em}
  } \\
  {\tt 1} &
  \multicolumn{2}{c}{
    \memoryaddrs{12em}
    \memory{3}{\unsealc}
    ~$\cdots$
    \MemoryLabel{-14em}{0.75em}{42}
    \MemoryLabel{-10em}{0.75em}{0}
    \MemoryLabel{-7em}{0.75em}{c}
    \vspace{.5em}
  } \\
  {\tt 10} &
  \memoryaddrs{20em}
  \memory{2}{\mainsealc}
  \memory{2}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{42}
  \MemoryLabel{-14em}{0.75em}{0}
  \MemoryLabel{-10em}{0.75em}{2}
  \MemoryLabel{-7em}{0.75em}{\bf d}
 &
  \memoryaddrs{20em}
  \memory{2}{\mainsealc}
  \memory{2}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{a'}
  \MemoryLabel{-14em}{0.75em}{b'}
  \MemoryLabel{-10em}{0.75em}{2}
  \MemoryLabel{-7em}{0.75em}{\bf d'}
  \\
  {\tt 11} &
  \memoryaddrs{20em}
  \memory{2}{\mainsealc}
  \memory{2}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{42}
  \MemoryLabel{-14em}{0.75em}{0}
  \MemoryLabel{-10em}{0.75em}{2}
  \MemoryLabel{-7em}{0.75em}{\bf 0}
  &
  \memoryaddrs{20em}
  \memory{2}{\mainsealc}
  \memory{2}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{a'}
  \MemoryLabel{-14em}{0.75em}{b'}
  \MemoryLabel{-10em}{0.75em}{2}
  \MemoryLabel{-7em}{0.75em}{\bf b'}
  \\
\end{tabular}

@@

\section{Extension to Cheri}
\label{sec:cheri}
